Quantum Hoare Types

### 2019-10-29 paper/paper.tex
General cleanup

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

%% Added by kartik
\usepackage[braket]{qcircuit}
\usepackage{mathpartir}

@@

%% Title information
\title[]{Quantum Hoare Types}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Author with single affiliation.
\author{Kartik Singhal}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{0000-0003-1132-269X}             %% \orcid is optional
\affiliation{
%  \position{Position1}
  \department{Department of Computer Science}              %% \department is recommended
  \institution{University of Chicago}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{United States}                    %% \country is recommended
}
\email{ks@cs.uchicago.edu}          %% \email is recommended

@@

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

@@

Previous work such as Proto-Quipper\cite{ross2015algebraic} and QWire\cite{paykin2018,rand2018a} utilize a linear type system and dependent types to enforce a small subset of semantic properties in the quantum setting such as the no-cloning theorem and whether a unitary operator is of the right dimension. These advances in quantum type systems although helpful, still fall short in encoding and enforcing even more useful properties that one would like to be able to express for the purpose of verification. Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Types in classical programming languages, we propose Quantum Hoare Types aimed at enabling both sound static type checking and formal verification of quantum programs.

@@

\begin{figure}
	\centerline{
		\Qcircuit @C=1em @R=.7em {
		\lstick{\ket{q_0}} & \gate{H} & \ctrl{1} & \qw \\
		\lstick{\ket{q_1}} & \qw      & \targ    & \qw
	}}
	\caption{Bell Pair Creation}
	\label{fig:bellpair}
\end{figure}

@@

\begin{figure}
\begin{tabular}{lrcl}
	\textit{Types} & $\tau, \sigma$ & ::= & $ \mathrm{qbit} \mid \tau \otimes \sigma \mid 1 \mid \Pi x{:}\tau.\sigma \mid \Psi. X. \{P\} x{:}\tau \{Q\}$ \\
	\textit{Assertions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{id}_\tau(M,N) \mid \mathrm{seleq}_\tau(H,M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid \\
		P \supset Q \mid \neg P \mid \forall h{:}\mathrm{heap}.P $
	\end{minipage}\\
	\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
	\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
	\textit{Intro terms} & $M, N, O$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ K \mid (\,) \mid \lambda x.\ M \mid \mathrm{do}\ E \mid \ket{0} \mid \ket{1} \mid (M, N) \mid M==N $
	\end{minipage}\\
	\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
		\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
	\end{minipage}\\
	\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
	\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
	\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
	\textit{Assertion context} & $\Gamma$& ::= & $\cdot \mid \Gamma, P $
\end{tabular}
	\caption{Syntax}
\label{fig:syntax}
\end{figure}

@@

Simple example to create a Bell Pair using QHTT commands:\\

@@

\hrulefill

From Xavier Leroy's OPLSS Slides:

Arithmetic expressions:

\[a ::= n \mid x \mid a1 + a2 \mid a1 - a2 \mid a1 \times a2\]

Boolean expressions:

\begin{align*}
	b ::=&\ \mathrm{true} \mid \mathrm{false} \mid a1 = a2 \mid a1 \le a2\\
	\mid&\ \mathrm{not}\ b \mid b1\ \mathrm{and}\ b2
\end{align*}

Commands (statements):

\begin{align*}
	c ::=&\ \mathrm{SKIP}	&\mathrm{(do\ nothing)}\\
	\mid&\ x ::= a	&\mathrm{(assignment)}\\
	\mid&\ 1; ; c2	&\mathrm{(sequence)}\\
	\mid&\ \mathrm{IFB}\ b\ \mathrm{THEN}\ c1\ \mathrm{ELSE}\ c2\ \mathrm{FI}	&\mathrm{(conditional)}\\
	\mid&\ \mathrm{WHILE}\ b\ \mathrm{DO}\ c\ \mathrm{END}	&\mathrm{(loop)}
\end{align*}

>>>>

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,nonacm]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

%% Added by kartik
\usepackage[braket]{qcircuit}
\usepackage{mathpartir}
\usepackage{xspace}

\newcommand{\sqir}{\ensuremath{s\textsc{qir}}\xspace}
\newcommand{\qwire}{\ensuremath{\mathcal{Q}\textsc{wire}}\xspace}
\newcommand{\denote}[1]{\llbracket #1 \rrbracket\xspace}

@@

%% Title information
\title[Quantum Hoare Types]{Quantum Hoare Types}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Author with single affiliation.
\author{Kartik Singhal}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{0000-0003-1132-269X}             %% \orcid is optional
\affiliation{
%  \position{Position1}
  \department{Department of Computer Science}              %% \department is recommended
  \institution{University of Chicago}            %% \institution is required
%  \streetaddress{Street1 Address1}
%  \city{City1}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{USA}                    %% \country is recommended
}
\email{ks@cs.uchicago.edu}          %% \email is recommended

@@

\ccsdesc[500]{Software and its engineering~General programming languages}
%% End of generated code

@@

Previous work such as Proto-Quipper\cite{ross2015algebraic} and \qwire\cite{paykin2018,rand2018a} utilize a linear type system and dependent types to enforce a small subset of semantic properties in the quantum setting such as the no-cloning theorem and whether a unitary operator is of the right dimension. These advances in quantum type systems although helpful, still fall short in encoding and enforcing even more useful properties that one would like to be able to express for the purpose of verification. Inspired by the use of Hoare triples in the verification of imperative programs and building on the idea of Hoare Types in classical programming languages, we propose Quantum Hoare Types aimed at enabling both sound static type checking and formal verification of quantum programs.

@@

\begin{figure}[hb]
	\centerline{
		\Qcircuit @C=1em @R=.7em {
		\lstick{\ket{q_0}} & \gate{H} & \ctrl{1} & \qw \\
		\lstick{\ket{q_1}} & \qw      & \targ    & \qw
	}}
	\caption{Bell Pair Creation}
	\label{fig:bellpair}
\end{figure}

@@

\begin{figure}[hb]
\begin{tabular}{lrcl}
	\textit{Types} & $\tau, \sigma$ & ::= & $ \mathrm{qbit} \mid \tau \otimes \sigma \mid 1 \mid \Pi x{:}\tau.\sigma \mid \Psi. X. \{P\} x{:}\tau \{Q\}$ \\
	\textit{Assertions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{id}_\tau(M,N) \mid \mathrm{seleq}_\tau(H,M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid \\
		P \supset Q \mid \neg P \mid \forall h{:}\mathrm{heap}.P $
	\end{minipage}\\
	\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
	\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
	\textit{Intro terms} & $M, N, O$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ K \mid (\,) \mid \lambda x.\ M \mid \mathrm{do}\ E \mid \ket{0} \mid \ket{1} \mid (M, N) \mid M==N $
	\end{minipage}\\
	\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
		\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
	\end{minipage}\\
	\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
	\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
	\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
	\textit{Assertion context} & $\Gamma$& ::= & $\cdot \mid \Gamma, P $
\end{tabular}
	\caption{Syntax}
\label{fig:syntax}
\end{figure}

@@

Simple example to create a Bell Pair using QHTT commands:

@@
